---
title: "NHL Win Probability Modeling (Portfolio Project)"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
options(repos = c(CRAN = "https://cran.r-project.org"))

library(tidyverse)
library(lmtest)
library(sandwich)
library(broom)
```

## 1) Load data

This repo does **not** redistribute large/raw datasets.  
To reproduce results, place `game_teams_stats.csv` in `data/` or run using the included sample.

```{r}
DATA_PATH_MAIN   <- "data/game_teams_stats.csv"
DATA_PATH_SAMPLE <- "data/game_teams_stats_sample.csv"

data_path <- if (file.exists(DATA_PATH_MAIN)) DATA_PATH_MAIN else DATA_PATH_SAMPLE
games_raw <- read_csv(data_path, show_col_types = FALSE)

glimpse(games_raw)
```

## 2) Clean + feature engineering (team-game → team vs opponent)

Each NHL game appears twice (home and away). We reshape to wide format and then reconstruct
a clean team-level dataset with opponent variables.

```{r}
games_clean <- games_raw %>%
  filter(!is.na(goals), !is.na(shots), !is.na(pim)) %>%
  distinct()

# Ensure each game has exactly 2 rows (home + away)
games_wide <- games_clean %>%
  filter(HoA %in% c("home","away")) %>%
  group_by(game_id) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  pivot_wider(
    names_from = HoA,
    values_from = c(team_id, won, settled_in, goals, shots, hits, pim,
                    powerPlayOpportunities, powerPlayGoals,
                    faceOffWinPercentage, giveaways, takeaways),
    names_sep = "_"
  )

# Reconstruct team perspective rows
games_home <- games_wide %>%
  transmute(
    game_id,
    team_id = team_id_home,
    home = 1L,
    win = as.integer(won_home),
    settled_in = settled_in_home,
    goals = goals_home,
    shots = shots_home,
    hits = hits_home,
    pim = pim_home,
    powerPlayOpportunities = powerPlayOpportunities_home,
    powerPlayGoals = powerPlayGoals_home,
    faceOffWinPercentage = faceOffWinPercentage_home,
    giveaways = giveaways_home,
    takeaways = takeaways_home,
    opp_goals = goals_away,
    opp_shots = shots_away,
    opp_pim = pim_away,
    opp_pp_opps = powerPlayOpportunities_away,
    opp_pp_goals = powerPlayGoals_away
  )

games_away <- games_wide %>%
  transmute(
    game_id,
    team_id = team_id_away,
    home = 0L,
    win = as.integer(won_away),
    settled_in = settled_in_away,
    goals = goals_away,
    shots = shots_away,
    hits = hits_away,
    pim = pim_away,
    powerPlayOpportunities = powerPlayOpportunities_away,
    powerPlayGoals = powerPlayGoals_away,
    faceOffWinPercentage = faceOffWinPercentage_away,
    giveaways = giveaways_away,
    takeaways = takeaways_away,
    opp_goals = goals_home,
    opp_shots = shots_home,
    opp_pim = pim_home,
    opp_pp_opps = powerPlayOpportunities_home,
    opp_pp_goals = powerPlayGoals_home
  )

games <- bind_rows(games_home, games_away) %>%
  mutate(
    gd = goals - opp_goals,
    shots_for = shots,
    shots_against = opp_shots,
    net_pp = powerPlayOpportunities - opp_pp_opps,
    pim_diff = pim - opp_pim
  )

summary(games$win)
```

## 3) Baseline logistic regression (win probability)

```{r}
model_baseline <- glm(
  win ~ pim + gd + shots_for + shots_against + home,
  data = games,
  family = binomial(link = "logit")
)

# Robust (HC0) standard errors
robust_test <- coeftest(model_baseline, vcov = vcovHC(model_baseline, type = "HC0"))
robust_test
```

### Odds ratios table (portfolio-friendly)

```{r}
or_table <- tidy(model_baseline) %>%
  mutate(
    OR = exp(estimate),
    OR_low = exp(estimate - 1.96 * std.error),
    OR_high = exp(estimate + 1.96 * std.error)
  )

or_table
```

```{r}
dir.create("outputs", showWarnings = FALSE)
write_csv(or_table, "outputs/baseline_odds_ratios.csv")
```

## 4) Predicted win probability vs. penalty minutes (controls fixed)

```{r}
dir.create("figures", showWarnings = FALSE)

pim_grid <- tibble(
  pim = seq(0, 40, by = 1),
  gd = 0,
  shots_for = median(games$shots_for, na.rm = TRUE),
  shots_against = median(games$shots_against, na.rm = TRUE),
  home = 0L
)

pim_grid$pred_win_prob <- predict(model_baseline, newdata = pim_grid, type = "response")

ggplot(pim_grid, aes(x = pim, y = pred_win_prob)) +
  geom_line() +
  labs(
    title = "Predicted Win Probability vs. Penalty Minutes",
    subtitle = "Controls fixed at: goal differential = 0, shots at median, away game",
    x = "Penalty minutes (PIM)",
    y = "Predicted P(win)"
  )
```

```{r}
ggsave("figures/pim_predicted_winprob.png", width = 8, height = 5, dpi = 200)
```

## 5) Optional: interaction (Home × PIM)

```{r}
model_home_int <- glm(
  win ~ pim * home + gd + shots_for + shots_against,
  data = games,
  family = binomial(link = "logit")
)

coeftest(model_home_int, vcov = vcovHC(model_home_int, type = "HC0"))
```


