---
title: "NHL Win Probability Modeling"
author: "Bangguo Jin"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(broom)
library(readxl)
library(gt)
library(gridExtra)
library(ggeffects)
```

## Overview

This cleaned version is designed for GitHub/portfolio use. It:

- uses a repo-friendly data path search
- keeps the author name as **Bangguo Jin**
- saves selected figures to a `figures/` folder
- keeps the original analysis logic while making the code easier to reuse

```{r helpers}
find_data_file <- function() {
  candidates <- c(
    "data/game_teams_stats.csv",
    "game_teams_stats.csv",
    "data/game_teams_stats_sample.csv",
    "game_teams_stats_sample.csv",
    "data/game_teams_stats.xlsx",
    "game_teams_stats.xlsx"
  )

  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) {
    stop(
      "No dataset found. Place one of these files in the project folder:\n",
      "- data/game_teams_stats.csv\n",
      "- game_teams_stats.csv\n",
      "- data/game_teams_stats_sample.csv\n",
      "- game_teams_stats_sample.csv\n",
      "- data/game_teams_stats.xlsx\n",
      "- game_teams_stats.xlsx"
    )
  }
  hit
}

load_nhl_data <- function(path) {
  if (grepl("\\.xlsx$", path, ignore.case = TRUE)) {
    df <- read_excel(path, sheet = 1)
    colnames(df) <- c(
      "game_id", "team_id", "HoA", "won", "settled_in", "head_coach",
      "goals", "shots", "hits", "pim", "powerPlayOpportunities",
      "powerPlayGoals", "faceOffWinPercentage", "giveaways",
      "takeaways", "blocked", "startRinkSide"
    )
  } else {
    df <- read_csv(path, show_col_types = FALSE)
  }

  df %>%
    mutate(
      game_id = as.character(game_id),
      team_id = as.numeric(team_id),
      HoA = as.character(HoA),
      won = as.logical(won),
      goals = as.numeric(goals),
      shots = as.numeric(shots),
      pim = as.numeric(pim),
      powerPlayOpportunities = as.numeric(powerPlayOpportunities),
      powerPlayGoals = as.numeric(powerPlayGoals)
    ) %>%
    filter(
      !is.na(team_id), !is.na(goals), !is.na(shots),
      !is.na(pim), !is.na(won)
    ) %>%
    distinct(game_id, team_id, .keep_all = TRUE)
}

dir.create("figures", showWarnings = FALSE, recursive = TRUE)
dir.create("outputs", showWarnings = FALSE, recursive = TRUE)

data_path <- find_data_file()
df <- load_nhl_data(data_path)

cat("Loaded file:", data_path, "\n")
cat("Rows after cleaning:", nrow(df), "\n")
```

## Question 1  
**Does higher total penalty minutes (PIM) per game decrease win probability after controlling for goal differential and shots?**

```{r q1-data}
df_wide_q1 <- df %>%
  select(game_id, team_id, HoA, won, goals, shots, pim) %>%
  pivot_wider(
    names_from = HoA,
    values_from = c(team_id, won, goals, shots, pim),
    names_glue = "{HoA}_{.value}"
  ) %>%
  filter(!is.na(home_team_id), !is.na(away_team_id))

analysis_q1 <- bind_rows(
  df_wide_q1 %>%
    transmute(
      game_id,
      location = "home",
      won = home_won,
      goal_diff = home_goals - away_goals,
      shots_self = home_shots,
      shots_opp = away_shots,
      pim_self = home_pim
    ),
  df_wide_q1 %>%
    transmute(
      game_id,
      location = "away",
      won = away_won,
      goal_diff = away_goals - home_goals,
      shots_self = away_shots,
      shots_opp = home_shots,
      pim_self = away_pim
    )
) %>%
  filter(complete.cases(.))

q1_model <- glm(
  won ~ pim_self + goal_diff + shots_self + shots_opp,
  data = analysis_q1,
  family = binomial
)

q1_results <- tidy(q1_model, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

q1_results %>%
  select(
    Term = term,
    `Odds Ratio` = estimate,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `p-value` = p.value
  ) %>%
  gt() %>%
  tab_header(title = "Question 1 Results")
```

```{r q1-figures}
# Figure A1: distribution of PIM
fig_a1 <- ggplot(analysis_q1, aes(x = pim_self)) +
  geom_histogram(binwidth = 7, fill = "#E69F00", color = NA) +
  labs(
    title = "Figure A1. Distribution of team penalty minutes per game",
    x = "Team penalty minutes in game (PIM)",
    y = "Number of team-games"
  ) +
  theme_minimal(base_size = 16)

# Figure 1A: observed win percentage by PIM
fig_1a_data <- analysis_q1 %>%
  group_by(pim_self) %>%
  summarise(
    win_pct = mean(as.numeric(won)),
    .groups = "drop"
  )

fig_1a <- ggplot(fig_1a_data, aes(x = pim_self, y = win_pct)) +
  geom_line(color = "#E69F00", linewidth = 1) +
  geom_point(color = "#E69F00", size = 2.8) +
  labs(
    title = "Figure 1A. Win percentage by team penalty minutes",
    x = "Team penalty minutes in game (PIM)",
    y = "Observed win percentage"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

# Figure 1B: predicted probability vs PIM
pred_q1 <- expand.grid(
  pim_self = seq(0, 40, by = 1),
  goal_diff = 0,
  shots_self = mean(analysis_q1$shots_self, na.rm = TRUE),
  shots_opp = mean(analysis_q1$shots_opp, na.rm = TRUE)
)

pred_q1$prob <- predict(q1_model, newdata = pred_q1, type = "response")

fig_1b <- ggplot(pred_q1, aes(x = pim_self, y = prob)) +
  geom_line(color = "#E69F00", linewidth = 1.2) +
  labs(
    title = "Figure 1B. Model-predicted win probability vs team PIM\n(Goal differential = 0, average shots)",
    x = "Team penalty minutes in game (PIM)",
    y = "Predicted probability of winning"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

fig_a1
fig_1a
fig_1b

ggsave("figures/pim_distribution.png", fig_a1, width = 11, height = 8, dpi = 200)
ggsave("figures/observed_win_percentage_by_pim.png", fig_1a, width = 11, height = 8, dpi = 200)
ggsave("figures/predicted_win_probability_vs_pim.png", fig_1b, width = 11, height = 8, dpi = 200)
```

## Question 2  
**Among teams that take more penalties than their opponent, does high power-play efficiency mitigate the negative effect?**

```{r q2-data}
q2_wide <- df %>%
  select(game_id, HoA, team_id, won, pim, powerPlayOpportunities, powerPlayGoals, goals) %>%
  pivot_wider(
    names_from = HoA,
    values_from = c(pim, powerPlayOpportunities, powerPlayGoals, won, goals, team_id),
    names_glue = "{HoA}_{.value}"
  ) %>%
  filter(
    !is.na(home_pim), !is.na(away_pim),
    !is.na(home_powerPlayOpportunities), !is.na(away_powerPlayOpportunities)
  )

analysis_q2 <- bind_rows(
  q2_wide %>%
    transmute(
      game_id,
      location = "home",
      won = home_won,
      goal_diff = home_goals - away_goals,
      pim_self = home_pim,
      pim_opp = away_pim,
      pp_opps_self = home_powerPlayOpportunities,
      pp_goals_self = home_powerPlayGoals
    ),
  q2_wide %>%
    transmute(
      game_id,
      location = "away",
      won = away_won,
      goal_diff = away_goals - home_goals,
      pim_self = away_pim,
      pim_opp = home_pim,
      pp_opps_self = away_powerPlayOpportunities,
      pp_goals_self = away_powerPlayGoals
    )
) %>%
  filter(pp_opps_self > 0) %>%
  mutate(
    more_penalties = pim_self > pim_opp,
    pp_conversion = pp_goals_self / pp_opps_self,
    high_pp_conv = pp_conversion >= 0.25
  ) %>%
  filter(complete.cases(.))

q2_subset <- analysis_q2 %>% filter(more_penalties)

q2_model <- glm(
  won ~ high_pp_conv + pim_self + goal_diff,
  data = q2_subset,
  family = binomial
)

q2_results <- tidy(q2_model, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

q2_results %>%
  select(
    Term = term,
    `Odds Ratio` = estimate,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `p-value` = p.value
  ) %>%
  gt() %>%
  tab_header(title = "Question 2 Results")
```

```{r q2-figures}
win_rate_summary <- q2_subset %>%
  group_by(high_pp_conv) %>%
  summarise(
    win_rate = mean(won),
    n_games = n(),
    .groups = "drop"
  ) %>%
  mutate(
    group = if_else(high_pp_conv, "High PP teams", "Other teams")
  )

fig_2a <- ggplot(win_rate_summary, aes(x = group, y = win_rate)) +
  geom_col(fill = "#E69F00", alpha = 0.95) +
  labs(
    title = "Figure 2A. Win percentage when taking more PIM than opponent\nby team power-play efficiency group",
    x = NULL,
    y = "Win percentage"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

pred_q2 <- expand.grid(
  pim_self = seq(0, 40, by = 1),
  high_pp_conv = c(FALSE, TRUE),
  goal_diff = 0
)

pred_q2$prob <- predict(q2_model, newdata = pred_q2, type = "response")

fig_2b <- ggplot(pred_q2, aes(x = pim_self, y = prob, color = high_pp_conv)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    values = c("#E69F00", "#56B4E9"),
    labels = c("Other teams", "High PP teams"),
    name = NULL
  ) +
  labs(
    title = "Figure 2B. Predicted win probability vs PIM\nwhen team has more PIM than opponent",
    x = "Team penalty minutes in game (PIM)",
    y = "Predicted probability of winning"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

fig_2a
fig_2b

ggsave("figures/high_pp_win_percentage_bar.png", fig_2a, width = 11, height = 8, dpi = 200)
ggsave("figures/high_pp_predicted_win_probability.png", fig_2b, width = 11, height = 8, dpi = 200)
```

## Question 3  
**Does the negative effect of PIM differ for home and away teams?**

```{r q3-data}
analysis_q3_base <- df %>%
  select(game_id, HoA, team_id, won, pim, goals, shots) %>%
  mutate(
    HoA = factor(HoA, levels = c("away", "home")),
    won = as.factor(won),
    across(c(pim, goals, shots), as.numeric)
  ) %>%
  filter(complete.cases(.))

q3_wide <- analysis_q3_base %>%
  select(game_id, HoA, goals, shots, pim, won) %>%
  pivot_wider(
    names_from = HoA,
    values_from = c(goals, shots, pim, won),
    names_glue = "{HoA}_{.value}"
  ) %>%
  filter(!is.na(home_goals), !is.na(away_goals))

analysis_q3 <- bind_rows(
  q3_wide %>%
    transmute(
      game_id,
      HoA = "home",
      won = as.factor(home_won),
      pim = home_pim,
      shots = home_shots,
      goal_diff = home_goals - away_goals
    ),
  q3_wide %>%
    transmute(
      game_id,
      HoA = "away",
      won = as.factor(away_won),
      pim = away_pim,
      shots = away_shots,
      goal_diff = away_goals - home_goals
    )
) %>%
  mutate(HoA = factor(HoA, levels = c("away", "home")))

q3_model_main <- glm(
  won ~ HoA + pim + goal_diff + shots,
  data = analysis_q3,
  family = binomial
)

q3_model_interaction <- glm(
  won ~ HoA * pim + goal_diff + shots,
  data = analysis_q3,
  family = binomial
)

q3_results <- tidy(q3_model_interaction, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

q3_results %>%
  select(
    Term = term,
    `Odds Ratio` = estimate,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `p-value` = p.value
  ) %>%
  gt() %>%
  tab_header(title = "Question 3 Results")
```

```{r q3-figures}
# Figure 3A: observed home vs away win percentage by PIM
fig_3a_data <- analysis_q3 %>%
  group_by(HoA, pim) %>%
  summarise(
    win_pct = mean(as.numeric(won) - 1),
    .groups = "drop"
  )

fig_3a <- ggplot(fig_3a_data, aes(x = pim, y = win_pct, color = HoA)) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = fig_3a_data %>% filter(HoA == "home"),
    size = 2.5
  ) +
  scale_color_manual(
    values = c("home" = "#E69F00", "away" = "#56B4E9"),
    labels = c("Away teams", "Home teams"),
    name = NULL
  ) +
  labs(
    title = "Figure 3A. Win percentage vs PIM for home and away teams",
    x = "Team penalty minutes in game (PIM)",
    y = "Observed win percentage"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

# Figure 3B: model-predicted home vs away win probability
pred_q3 <- ggpredict(q3_model_interaction, terms = c("pim [0:40 by=1]", "HoA")) %>%
  as_tibble()

fig_3b <- ggplot(pred_q3, aes(x = x, y = predicted, color = group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    values = c("home" = "#E69F00", "away" = "#56B4E9"),
    labels = c("Away teams", "Home teams"),
    name = NULL
  ) +
  labs(
    title = "Figure 3B. Predicted win probability vs PIM for home and away teams",
    x = "Team penalty minutes in game (PIM)",
    y = "Predicted probability of winning"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 16)

fig_3a
fig_3b

ggsave("figures/home_away_observed_win_percentage.png", fig_3a, width = 11, height = 8, dpi = 200)
ggsave("figures/home_away_predicted_win_probability.png", fig_3b, width = 11, height = 8, dpi = 200)
```

## Saved figure files

```{r saved-files}
list.files("figures")
```
